---
title: "Introducing R"
subtitle: "Data Type, Data Structure and Operation (v2.0)"
author: "Chia-hung Tsai (Election Study Center, NCCU)"
date: 'Wednesday, July 11, 2018, 5:10-6:30 p.m.'
output: 
  html_document: 
    toc: true # table of content true
    toc_depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: true  ## if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
    fig_caption: true
    #pandoc_args: ["-F","pandoc-crossref"]
    toc_float:
      collapsed: false
      smooth_scroll: false
---
<style>
body {
    background-color: #EEE8AA;
    font-size: 22px;
    color: #171717;
    font-family: Times, "BiauKai", sans-serif;
    line-height: 1.8;
}
blue {font-family: courier, lucida;  
      color: blue}
fr  {font-family: georgia; 
      color: firebrick}
</style>

```{r setup, include=FALSE, echo=FALSE}
library(knitr)
library(dplyr)
library(kableExtra)
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE, digits = 3)
options(knitr.table.format = "html", digits = 3)
```

# Goal of this course
We will introduce the data type and data structure of `R`. For example, we use `c()` to denote a set of numbers and strings.
```{r echo=TRUE}
A<-c("Taipei City","New Taipei City", "Taoyuan City", "Taichung City","Tainan City","Kaohsiung City")
print(A)
B<-c(0,1,2,3,4,5,6,7,8,9)
print(B)
```


---

# Overview

- One dimension
    - vector
    - factor
- Two dimensions
    - matrix
    - data frame
- Multiple dimensions
    - array
    - list  
    - table

<hr>
# What is `R`?
`R` is a free software for data analysis and statistical computing. It can be download from [here](https://cran.r-project.org). It has a base package and allows users to download packages from a specific repository or the "Comprehensive R Archive Network" (CRAN).  
Please make sure that you download the right version compatible with your system.     
`RStudio` is an IDE that expands the functions of `R`. For example, you can embed `R` codes in a html or $\LaTeX$ document. You can also import different kinds of datasets easily. Notice that it cannot run without `R`.    

# Data Type
## numeric
Numeric data can be numeric or integer. For example:

```{r echo=TRUE}
x<-c(2, 4, 6, 8); x
class(x)
```

The function <fr>class()</fr> returns the data type.

We can also use scientific notations to present the numeric data:

```{r echo=TRUE}
y=c(1.1e+06); y
class(y)
```


Integer is a subset of numbers. In a 32-bit operation system, the maximum integer is 2147483647，far smaller than the maximum of number. Every number is composed of significan ($+\-$), mantissa and exponent to be computed, in particular the number with a decimal. For example:
```{r echo=TRUE}
u<-as.integer(c(4)); class(u)
```

Another example:   
```{r}
a=1.356e+3
is.numeric(a)
is.integer(a)
options(digits=20)
pi
is.numeric(pi); is.integer(pi)
```

Integer can be added, substracted, multiplied, and divided like number.

```{r}
J<-c(3, 6, 33)
J<-as.integer(J)
J-1; J*2; J/3
```

## character

Charaters are easy to understand, espcially for data visualization. For example: 
```{r echo=TRUE}
library(lattice)
s77<-data.frame(state.x77)
s77<-s77[order(s77$Population,decreasing=T),]
dotchart(s77$Population,
labels=row.names(s77), pch=16,cex=.7, xlab='Population')
```

Here we transform the matrix **state.x77** to be a data frame, and then we draw a <fr>dotchart</fr>, a function from `lattice` package.  

numbers can be treated as characters:
```{r echo=TRUE}
char1<-c("1","2", "Do", "Re", "Mi"); char1
```

Can we convert characters to numbers? No. But we can convert it to factor, then converting it to numbers.
```{r}
LETTERS
LETTERS.f <- as.factor(LETTERS)
as.numeric(LETTERS.f)
```

Please read the **councilor** data. Notice that we can encode the data if it is not UTF-8.

```{r}
library(foreign)
df <- read.csv('councilor.csv', header=T, fileEncoding = 'BIG5', 
               colClasses = "character")
df
as.numeric(df$unit)
```

To make the conversion easier, we change the arguments <blue>colClasses="factor"</blue> in the <fr>read.csv()</fr> funcation.
```{r}
library(foreign)
df <- read.csv('councilor.csv', header=T, fileEncoding = 'BIG5', 
               colClasses = "factor")
df
as.numeric(df$unit)
```

## logic

<li>Data can be logic as True or False. We use it to filter data. For example, let's create a logic variable **ok**: </li>
```{r echo=TRUE}
s77$ok<-s77$Income>5000; s77$ok
```

<li>We can select cases of a data frame according to this variable:</li>

```{r echo=TRUE}
options(digits=3)
s77[s77$ok, ]
```

Please try to run the following codes:    

```markdown
library(car)
head(Duncan)
income.hi<-Duncan$income>70  
Duncan[income.hi, ]
```
How many "bc" are left in your dataset?

## date

<li>Date is a special type of data. First of all, we can show what date is today. </li>
```markdown
Sys.Date()
```
<li>If we have a vector of characters, we can tranform it to date according to the original format. We can transform it to other date format.</li>

```{r echo=TRUE}
v<-c("2/27/2018", "6/26/2018", "12/31/2018"); class(v)
v.date<-as.Date(v, format='%m/%d/%Y'); class(v.date)
v.date
v.date <-format(v.date, "%b. %d, %Y"); v.date
```

Or
```{r echo=TRUE}
v<-c("", "6/26/2018", "12/31/2018")
as.Date(v, format='%m/%d/%Y')
```
<li>We can also know certain dates after certain days pass from the origin date. </li>
```{r echo=T}
S <- c(50, 100)
as.Date(S, origin="2018-01-01")
```

<li>So what is the strength of date? Can we stick with character or numeric? It depends. </li>
<li>For example, we create the following dataset with <fr>tibble()</fr>:</li>
```{r}
library(dplyr)
df <- tibble(date=c("2016", "2017", "2018"),
                    students=c(20, 22, 18),
                    teachers=c(12, 13, 20))
df
```
<li>Then let's draw a multiple-line chart to visualize the trends of students and teachers. In order to do so, we use <fr>melt()</fr> to create a long table according to the index variable, **date**. </li>

```{r}
library(reshape2); library(ggplot2)
df2 <- melt(df, id.vars="date", variable.name="Group")
df2$Group <-as.factor(df2$Group)
df2
ggplot(df2, aes(x=date, y=value, col=Group)) + 
     geom_line( size=1) +
     geom_point(shape=6, size=3)           
```

<li>The figure looks okay, but there are no lines connecting the dots. Let's convert the type of "date" to date type.</li>

```{r}
library(reshape2); library(ggplot2)
df2 <- melt(df, id.vars="date", variable.name="Group")
df2$Group <-as.factor(df2$Group)
df2$date <- as.Date(df2$date, format="%Y")
df2
ggplot(df2, aes(x=date, y=value, col=Group)) + 
     geom_line( size=1) +
     geom_point(shape=16, size=3)           
```

<li>We can try another dataset as follows.</li>
```{r}
library(dplyr)
df <- tibble(date=c("2018/07/11", "2018/07/12", "2018/07/13"),
                    students=c(20, 22, 18),
                    teachers=c(2, 3, 4))
df             
library(reshape2); library(ggplot2)
df2 <- melt(df, id.vars="date", variable.name="Group")
df2$Group <-as.factor(df2$Group)
df2$date <- as.Date(df2$date, format="%Y/%m/%d")
df2
ggplot(df2, aes(x=date, y=value, col=Group)) + 
     geom_line( size=1) +
     geom_point(shape=16, size=3) +
     scale_x_date(date_labels = "%Y/%m/%d")
```

<li>Here we use <fr>scale_x_date(date_labels = "%Y/%m/%d")</fr> to specify the date format of x-axis.</li>
### Difference in date

<li>We can calculate the difference between two dates. For example, we want to know the difference of Xi and Tsai's age: </li>
```{r echo=T}
xi<-"1953-06-15" #Xi's birthday
tsai<-"1956-08-31" #Tsai's birthday
```

<li>We use <fr>difftime()</fr> to calculate the difference:</li>
```{r echo=T}
as.Date(c(xi,tsai))
difftime(tsai, xi)
```


### Formats of Date

<li>Here are some formats of date: </li>

|symbol|meaing|example|
|--------|----|----|
|%d| date as number|01-31 |
|%a| abbreviated weekday|Mon |
|%A| unabbreviated weekday| Monday|
|%m| month as number|01-12 |
|%b| abbreviate month| Jan |
|%B| unabbreviate month|January |
|%y| 2-digit year| 18 |
|%Y| 4-digit year| 2018 |

<li><fr>format()</fr> can transform date to other formats. </li>
```{r echo=TRUE}
Today<-Sys.Date(); Today
to_day<-format(Today, format='%Y-%b-%d'); to_day
to_day<-format(Today, format='%m %d (%a), %y'); to_day
```

---

# Data structure
## one-dimension
### vector  

<li>Vector is the most often used data structure. </li>
```{r echo=TRUE}
example<-c(0,1,2,3,4)
print(example)
```

<li>We can reverse the arrow.</li>

```{r echo=TRUE}
c(2,4,6,8)->A
```

<li>There are upper case and lower case of English letters.</li>
```{r echo=TRUE}
c(letters)
c(LETTERS)
```

<li>We can directly do arithmetic within a vector. </li>
```{r echo=TRUE}
j<-c(2*2,2*9, 10-2, sqrt(25)); j
```

<li>We can multiply, divide, add, or substract a vector by certain number.</li>
```{r echo=TRUE}
R<-c(30, 40, 50); A <- R/5; A; A^2 
```
### Attribute
<li>We can use <fr>attr()</fr> function to store metadata about the object. Attributes can be thought of as a named list (with unique names). Attributes can be accessed with attr() or all at once (as a list) with attributes().</li>
```{r}

attr(mtcars$wt, "myattribute") <-"Weights"
attr(mtcars$wt, "mylabel") <-"Weights of Cars"
attr(mtcars$wt, "myattribute")
attr(mtcars$wt, "mylabel")
attributes(mtcars$wt)

```

<li>We can give names to elements in a vector.</li>
```{r}
U <- c(60, 70, 75, 82)
attr(U, "myattribute")<-"humanity"
names(U) <-c("Taichung", "Chiayi", "Tainan", "Kaohsiung")
U
barchart(U)
```


<li>We can combine vectors with vectors.</li>
```{r}
RA <- c(R, A)
RA
```

<li>If two vectors are of unequal length, the shorter one will be recycled in order to match the longer vector. </li>
```{r}
M1 <- c(1:3); M1
M2 <- rep(10, 4); M2
M1 + M2
```

### factor
Data can have categories, which look like characters.
```{r echo=TRUE}
library(ggplot2)
class(diamonds$cut)
table(diamonds$cut)
```

<li>In this example, **cut** is an ordered factor. We can try to do it on our own. </li>

```{r}
H<-c("Hi", "Lo", "Lo", "Middle", "Middle", "Middle")
table(H)

H.o <- ordered(H, levels=c("Lo", "middle", "Hi"))
table(H.o)
```

<li>Factor is useful because it has characters that we can read in tabulation.</li>
```{r echo=TRUE}
table(diamonds$cut, diamonds$color)
```

<li>It is also useful when we vizualize our data.</li>
```{r echo=TRUE}
library(car)
library(lattice)
plot( Chile$vote ~ Chile$sex, xlab="Sex", ylab="Vote")
```

#### Conversion of factor to numeric
<li>We can convert factor to numeric data.</li>
```{r echo=TRUE}
library(car)
data(Chile)
class(Chile$region)
table(Chile$region)
y<-as.numeric(Chile$region)
table(y)
```

<li>We can also create a numeric variable and replace factor into numeric.</li>
```{r echo=T}
Chile$gender[Chile$sex=="F"]<-1
Chile$gender[Chile$sex=="M"]<-2
table(Chile$gender)
table(Chile$sex)
```

<li>Please try to convert the **cut** variable from factor to numeric. </li>


## Two-dimension
### matrix

<li>We can create a matrix to represent two dimensions of data. For example, we can create a $3\times 3$ matrix like that:</li>
```{r echo=TRUE}
m<-matrix(c(1:9), nrow=3, ncol=3); m

n<-matrix(c(1:6), nrow=3, ncol=2); n

```
<li>We can multiply the two matrixes if the number of columns of the first matrixt equals to the number of row of the second matrix:</li>
```{r echo=TRUE}
m%*%n
```

<li>We can obtain the diagnol of matrix with <fr>diag()</fr> function:</li>
```{r echo=T}
diag(m)
diag(n)
```

<li>Sometimes we need to transpose the matrix: </li>
```{r echo=T}
t(m)
```

<li>We can replace one of the elements in a matrix with other numbers or strings by specifing the two dimensions:
```{r echo=T}
m
m1<-m
m1[2,3]<-0
m1[3,]<-99
m1
```

<li>We can set up different dimensions of a same matrix: </li>
```{r}
a <- matrix(c(1:20), nrow=2, ncol=10)
b <- matrix(c(1:20), nrow=5, ncol=4)
a; b

```

### data frame
<li>Data frame is like matrix. It can combines vectors and assign names to each of them.</li>
```{r echo=TRUE}
options(digits=4)
R1<-c(170, 175, 166, 172, 165, 157, 167, 167, 
        156, 160)
R2<-c("F","M","M","M","F","F","F","F","M","F")
R3<-R1/10 + 42

R123<-data.frame(height=R1,gender=R2,weight=R3); R123
```
<li>It is shown that **gender** is a factor. We can ask `R` not to convert it.</li>
```{r}
class(R123$gender)
R123<-data.frame(height=R1,gender=R2,weight=R3, stringsAsFactors = F)
str(R123)
```

<li>Each column must have same numbers of elements, and each row must have the same number of columns.</li>


-  <fr>nrow()</fr>：number of rows   
-  <fr>ncol()</fr>：number of columns  
-  <fr>dim()</fr>：number of rows and columns     
-  <fr>head()</fr>：the first six rows  
- <fr>str()</fr>: information of all variables  

<li>If we want to know the number of observations in <blue>AMSsurvey</li>：</li>
```{r echo=TRUE}
library(car)
nrow(AMSsurvey)
```

<li>If we want to change the names of variables of a data frame, we can use <fr>colnames()</fr> function:</li>
```{r echo=T}
newsurvey <- AMSsurvey
colnames(newsurvey)<-c("v1","v2","v3", "v4", "v5"); head(newsurvey)
```

<hr>

## Multi-dimension
### array
<li>Array can contain more than one matrix. An one-matrixt array is a matrix.</li>
```{r echo=TRUE}
Array1 <- array(1:12, dim = c(2, 6, 1)); Array1
```

<li>We can create an array that contains two $2\times 4$ matries</li>
```{r echo=TRUE}
Array2 <- array(16:1, dim = c(2, 4, 2)); Array2
```

<li>We can subset the array by specifying the one of the three dimesnions:</li>
```{r echo=TRUE}
A12<-Array2[,,2]; A12
A111 <- Array2[1, 1, 1]; A111
```

### list
<li>If we have vectors of unequal length, we can use list to store the data.</li>
```{r echo=T}
options(digits=4)
birthday <- c("1981/03/15", "1983/04/20", "1984/01/18")
listA<-list(R123, m, birthday); listA
listA[[3]]
```

<li>To organize the list, we can assign names to vectors, arrays, or matrix respectively. </li>
```{r echo=TRUE}
options(digits=4)
listB<-list(data=R123, vec=m, char=birthday); 
listB[["data"]]

```

```markdown
Please combine `c('a','b','c')`、`c(1,2,3,4)`以及`c('2018-01-01', '2018-04-04', '2018-04-05', '2018-06-18', '2018-10-10')` as a list.  
```
### table
<blue>Titanic</blue> is a table. It contains four arrays.
```{r echo=TRUE}
class(Titanic); Titanic

Titanic[, , 1, 1]
```

<li>We can use <fr>prop.table()</fr> to know the conditional probability of a table and <fr>margin.table()</fr> can return the marginal probability. </li>
```{r echo=T}
options(digits=4)
g<-Titanic[ , , 2, 2]; g
prop.table(g, 1)
prop.table(g, 2)
margin.table(g,1)
margin.table(g,2)
```
<br>

# Basic Operation   
## Operation of Vector   
<li>Because vector can represent the coordinates, so the order of elements is important, especially when we do arithmetic of two vectors.</li>
<li>First of all, we do arithmetic between a vector and a scalar.</li>
```{r echo=T}
X<-c(10,20,30,40,50,60); Sca<-10
X+Sca
X/Sca
```


<li>Then we try the operation of two vectors:</li>
```{r echo=T}
Y<-c(5,10,6,8,25,6)
X/Y; X*Y
```

## Mathmatical functions 
Please see more details [here](http://www.dummies.com/programming/r/how-to-use-mathematical-functions-in-r/)     

- sqrt(x): square root    
- exp(x): exponential of x    
- log(x, y): natural logarithm of x if y is not specified    
```{r}
a=6
exp(a); log(a)
log(exp(a)); exp(log(a))
```
- factorial(x): factorial of x. For example   
```{r}
1*2*3
factorial(3)
```
- abs(x): absolute value of x

## Rounding of numbers
Please see more details [here](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Round.html)      

- `round`：rounds the values in its first argument to the specified number of decimal places 
- `floor`：takes a single numeric argument x and returns a numeric vector containing the largest integers not greater than the corresponding elements of x.
- `ceiling`： takes a single numeric argument x and returns a numeric vector containing the smallest integers not less than the corresponding elements of x    

```{r echo=TRUE}
a1<-c(2.54, 3.111, 10.999)
round(a1, digits=2)
floor(a1)
ceiling(a1)
```

---

# Exercises

1. $\text{log}(\frac{14}{5})=$?  
```{r include=F}
log(14)-log(5)
```
2. $1\times 2\times 3\times , \dots ,\times 8=$?
```{r include=F}
factorial(8)
```
3. About **Titanic** data, please show which coach that the survived children were taking.
```{r include=F}
options(digits=4)
t<-Titanic[ , , 1, 2]; t
margin.table(t, 1)

```
4. Please use <fr>weekdays()</fr> function to show what day it is today.
```{r include=FALSE}
TODAY<-as.Date('2018-07-11', format='%Y-%m-%d')
weekdays(TODAY)
```

5. How many observations that has **mpg** greater or equal to 21 in `mtcars`？
```{r include=FALSE }
library(car)
knitr::kable(mtcars)
mtcars$Mpg<-mtcars$mpg
ok<-mtcars$Mpg>=21
mtcars$Mpg[ok==T]<-'hi'
mtcars$Mpg[ok==F]<-'lo'
table(mtcars$Mpg)
```

6. Please analyze if admission is related to gender in A department in <blue>UCBAdmissions</blue>. 
```{r include=FALSE, echo=F}
A <- UCBAdmissions[, , 1]; A
prop.table(A, 2)
#conclusion: not strongly related.
```
7. Please count the number of letters in English. 
```{r include=FALSE}
n<-length(c(LETTERS)); n
```
8. Please transform today's temperature to Fahrenheit.
```{r include=FALSE}
a<-28
F.a<-a*(9/5)+32; F.a
```

9. Please create a matrix with a diagnol (1,1,1)
```{r include=FALSE}
a<-matrix(c(1:9), nrow=3, ncol=3)
diag(a)<-c(1,1,1); a
```
10. Please count the number of days between when you are doing assignment and Jan. 1st.
```{r include=FALSE}
a<-Sys.Date(); b<-as.Date("2018-01-01")
difftime(a, b, units=c("days"))
```